Type theories as quotient inductive-recursive types

### 2025-02-21 tex/main.tex
cosmetic changes
final (?) pass Sections 1, 2

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules~\cite{Cockx2021,Cockx2020} after an inductive definition.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition in type theory.
%
Any method that reduces the friction introduced by equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

@@

In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
The prototypical example is Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code
\[
  \pi : (a : U) \to (T\,a \to U) \to U
  \quad\text{with}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
\]
for dependent function types makes sense only because we have access to the function $T$.
%

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given below
%\begin{figure}[b]
\begin{mathpar}
  \Jnil : \implicit{A}\; \JList{A} \and
  \Jsing{\blank} : \implicit{A}\; A \to \JList{A} \and
  \Jconcat{\blank}{\blank} : \implicit{A}\; \JList{A} \to \JList{A} \to \JList{A} \and
  \mathsf{idr} : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \and
  \mathsf{idl} : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \and
  \mathsf{assoc} : \implicit{\ell_0, \ell_1, \ell_2}\; \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{mathpar}
%\begin{minipage}{.6\textwidth}
%\begin{alignat*}{3}
%  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
%  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
%  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
%\end{alignat*}
%\end{minipage}
%\begin{minipage}{.4\textwidth}
%\begin{alignat*}{3}
%\end{alignat*}
%\end{minipage}
%\caption{The quotient inductive type of join lists}
%\label{fig:QIT-join-list}
%\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list and replaces equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'.
\]
This constructor states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
%In this work, we will consider quotient inductive-inductive-recursive types, generalising QIITs and QITs.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
In a proof assistant such as \Agda, functions are defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following a \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
%This elaboration offers conventional notation, while making computational behaviour opaque.
% and favouring asymmetric definitions.
For example, the two definitions of addition $+$ and $+'$
%
%\begin{figure}[b]%
\begin{minipage}{.48\textwidth}
  \begin{alignat*}{5}
%    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
%  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
%\caption{Two seemingly different definitions of addition}
%\label{fig:additions}
%\end{figure}
are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict while clauses of $+$ are not necessarily strict.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

@@

\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{5}
  \emptyctx     & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{5}
\subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
  \UU         & : \implicit{\Gamma}\; && \Ty\; \Gamma \\
  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma \\
  \Pi         & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
\end{alignat*}

>>>>

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We thank Hsiang-Shang Ko for useful discussions.}%optional

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules~\cite{Cockx2021,Cockx2020} after an inductive definition.
%
As the host theory is assumed to satisfy uniqueness of identity proofs and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition in type theory.
%
Any method that reduces the friction introduced by equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

@@

In contrast, an inductive-recursive type (IRT), introduced by Dybjer~\cite{Dybjer2000} and studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}, is an inductive type $A$ defined \emph{simultaneously} with a recursively defined function $F : A \to D$ to some fixed (possibly large) type $D$.
The prototypical example is a Tarski universe: a type $U$ of codes for small types is defined with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code
\[
  \pi : (a : U) \to (T\,a \to U) \to U
  \quad\text{with decoding}\quad T(\pi(a, b)) \defeq (x : T\,a) \to T\,(b\, x)
\]
for dependent function types makes sense only because we have access to the function $T$.
%

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given as follows:
%\begin{figure}[b]
\begin{mathpar}
  \Jnil : \implicit{A}\; \JList{A} \and
  \Jsing{\blank} : \implicit{A}\; A \to \JList{A} \and
  \Jconcat{\blank}{\blank} : \implicit{A}\; \JList{A} \to \JList{A} \to \JList{A} \and
  \mathsf{idr} : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \and
  \mathsf{idl} : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \and
  \mathsf{assoc} : \implicit{\ell_0, \ell_1, \ell_2}\; \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{mathpar}
%\begin{minipage}{.6\textwidth}
%\begin{alignat*}{3}
%  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
%  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
%  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
%\end{alignat*}
%\end{minipage}
%\begin{minipage}{.4\textwidth}
%\begin{alignat*}{3}
%\end{alignat*}
%\end{minipage}
%\caption{The quotient inductive type of join lists}
%\label{fig:QIT-join-list}
%\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi, and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursive function $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list, and we replace the equality constructors $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'.
\]
This constructor states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.
%In this work, we will consider quotient inductive-inductive-recursive types, generalising QIITs and QITs.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
In a proof assistant such as \Agda, functions are defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following a \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
%This elaboration offers conventional notation, while making computational behaviour opaque.
% and favouring asymmetric definitions.
For example, the two definitions of addition $+$ and $+'$
%
%\begin{figure}[b]%
\begin{minipage}{.48\textwidth}
  \begin{alignat*}{5}
%    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
%  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && = m  \\
  (\suc\;n)      & +' 0         && = \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && = \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
%\caption{Two seemingly different definitions of addition}
%\label{fig:additions}
%\end{figure}
are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict while clauses of $+$ are not necessarily strict.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

@@

\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{5}
  \emptyctx     & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{5}
\subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
  \UU         & : \implicit{\Gamma}\; && \Ty\; \Gamma \\
  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;\UU \to \Ty\;\Gamma \\
  \Pi         & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
\end{alignat*}
